---
title: "ATAC and RNA exploration"
author: "Koen Van den Berge"
date: "7/22/2020"
output: 
  html_document:
    toc: true
    toc_float: true
---


```{r pkg}
here::set_here()
suppressPackageStartupMessages({
  library(slingshot)
  library(SingleCellExperiment)
  library(cowplot)
  library(rgl)
  library(clusterExperiment)
  library(RColorBrewer)
  library(ggplot2)
  library(pheatmap)
  library(wesanderson)
  library(gridExtra)
  library(irlba)
  library(tidyverse)
})
```

# Import data

```{r loadData}
sds <- readRDS("../data/finalTrajectory/sling.rds")
counts <- readRDS("../data/finalTrajectory/counts_noResp_noMV.rds")
# counts <- round(counts)
sce <- readRDS("../data/finalTrajectory/sce_tradeSeq20200904.rds")
load("../data/ALL_TF.Rda")
clDatta <- readRDS("../data/finalTrajectory/dattaCl_noResp_noMV.rds")
timePoint <- readRDS("../data/finalTrajectory/timePoint_noResp_noMV.rds")
```

# Select activated and renewed HBC clusters

I will select 

 - activated HBCs as cells sampled in the 24h timepoint that are in the starting cluster
 - renewed ('resting') HBCs as cells sampled in the 14D timepoint that are in the HBC cluster

```{r}
cols <- c("#E41A1C", "#377EB8", "#4DAF4A", "#984EA3", "#FF7F00", "#FFFF33", "#A65628", "#F781BF", "#999999")
view3d( theta = 10, phi = 10)
# Datta labels
plot3d(reducedDim(sds), aspect = 'iso', col=cols[clDatta], alpha=.6, xlab="UMAP1", ylab="UMAP2", zlab="UMAP3", box=FALSE, top=TRUE)
plot3d(sds, add=TRUE, lwd=3, col="black")

# timepoints
plot3d(reducedDim(sds), aspect = 'iso', col=cols[timePoint], alpha=.6, xlab="UMAP1", ylab="UMAP2", zlab="UMAP3", box=FALSE, top=TRUE)
plot3d(sds, add=TRUE, lwd=3, col="black")

# cluster labels used for TI
clSds <- apply(slingClusterLabels(sds), 1, which.max)
plot3d(reducedDim(sds), aspect = 'iso', col=cols[clSds], alpha=.6, xlab="UMAP1", ylab="UMAP2", zlab="UMAP3", box=FALSE, top=TRUE)
plot3d(sds, add=TRUE, lwd=3, col="black")


actHBCId <- which(clSds == 8 & timePoint == "24HPI")
restHBCId <- which(clDatta == "HBC" & timePoint == "14DPI")
grpHBC <- rep(NA, length(c(actHBCId, restHBCId)))
grpHBC[1:length(actHBCId)] <- "activated"
grpHBC[is.na(grpHBC)] <- "resting"
grpHBC <- factor(grpHBC)
countHBC <- counts[, c(actHBCId, restHBCId)]
```

# Dimensionality reduction

```{r}
library(scran)
library(uwot)
library(scater)
logFracs <- log(sweep(countHBC+1e-5, 2, colSums(countHBC), "/"))
gv <- modelGeneVar(countHBC)
hvg <- getTopHVGs(gv, n=1000)
pc10 <- irlba::irlba(logFracs[hvg,], nv=10)
um10 <- uwot::umap(pc10$v %*% diag(pc10$d))

plot(um10, col=alpha(c("orange", "darkseagreen3")[as.numeric(grpHBC)], .75), pch=16, cex=1/2, bty='l')
legend("topright", c("activated", "resting"), pch=16, col=c("orange", "darkseagreen3"), bty="n")
```

# Batch correction with Harmony

```{r}
library(harmony)
harmony_embeddings <- HarmonyMatrix(logFracs, meta_data=grpHBC, do_pca=TRUE)
umHarmony <- uwot::umap(harmony_embeddings)

plot(umHarmony, col=alpha(c("orange", "darkseagreen3")[as.numeric(grpHBC)], .75), pch=16, cex=1/2, bty='l')
legend("topright", c("activated", "resting"), pch=16, col=c("orange", "darkseagreen3"), bty="n")
```

# Integration: Processing and dimensionality reduction of each dataset separately

```{r}
library(Signac)
library(Seurat)
library(patchwork)
library(ArchR)
set.seed(1234)
addArchRGenome("mm10")


# scATAC-seq importing and preprocessing
# from 20200610_scATAC_injury_archr_samples_pairwiseHBC_v3.Rmd
archProj <- loadArchRProject("../data/scATAC/archR_samples_pairwise_v2")
clHBCMerged <- archProj$clHBCMerged
plotEmbedding(ArchRProj = archProj, colorBy = "cellColData", 
              name = "clHBCMerged", embedding = "UMAP_cor")
archrUmap <- ArchR::getEmbedding(archProj, embedding = "UMAP_cor")
colnames(archrUmap) <- c("UMAP1", "UMAP2")
plotEmbedding(ArchRProj = archProj, colorBy = "cellColData", 
              name = "treat", embedding = "UMAP_cor")
clHBCMerged2 <- as.character(getCellColData(archProj, "clHBCMerged")$clHBCMerged)
clHBCMerged2[clHBCMerged2 == "C1_Inj"] <- "Activated"
clHBCMerged2[clHBCMerged2 == "C2_Hybrid"] <- "Hybrid"
clHBCMerged2[clHBCMerged2 == "C3_UI"] <- "Resting"
archrUmap$clHBCMerged2 <- factor(clHBCMerged2)
archrUmap$treatment <- getCellColData(archProj, "treat")$treat
peaks <- archProj@peakSet
names(peaks) <- paste0("peak",1:length(peaks))
atacPeakMat <- readRDS("../data/scATAC/peakMatArchrHbc_v2.rds")
rownames(atacPeakMat) <- names(peaks)
geneActivity <- readRDS("../data/scATAC/geneMatArchrHbc.rds")
atac <- Seurat::CreateSeuratObject(counts = atacPeakMat,
                           assay = "ATAC",
                           project = "10x_ATAC")
atac[["ACTIVITY"]] <- CreateAssayObject(counts = geneActivity)
atac$tech <- "atac"
atac$treatment <- archProj@cellColData$treat
atac$clHBCMerged <- clHBCMerged
# process gene activity
DefaultAssay(atac) <- "ACTIVITY"
atac <- FindVariableFeatures(atac)
atac <- NormalizeData(atac)
atac <- ScaleData(atac)
# process peak matrix
DefaultAssay(atac) <- "ATAC"
# VariableFeatures(atac) <- names(which(Matrix::rowSums(atac) > 50))
# atac <- RunLSI(atac, n = 30, scale.max = NULL)
# atac <- RunUMAP(atac, reduction = "lsi", dims = 1:30)
atac <- FindTopFeatures(atac, min.cutoff = 'q0')
# Seurat 3
atac <- RunLSI(atac, n = 30) 
# Seurat 4
#atac <- RunTFIDF(atac)
#atac <- RunSVD(atac, n = 30, reduction.key = "lsi_", reduction.name = "lsi")
# drop first dim as often correlated with seq depth
atac <- RunUMAP(object = atac, reduction = 'lsi', dims = 2:30)

# scRNA-seq import
rna <- CreateSeuratObject(counts=countHBC)
rna$tech <- '10x'
rna$celltype <- grpHBC
rna <- FindVariableFeatures(rna, selection.method = "vst", nfeatures = 2000)
rna <- ScaleData(rna)
rna <- RunPCA(rna, npcs=20)
rna <- RunUMAP(rna, dims = 1:20, min_dist=.1)


p1 <- DimPlot(atac, reduction = "umap", group.by="treatment", label=TRUE) + 
  NoLegend() + 
  ggtitle("scATAC-seq")
p2 <- DimPlot(rna, group.by = "celltype", label = TRUE, repel = TRUE) + NoLegend() + ggtitle("scRNA-seq")
p1 + p2


# Markers for activated/resting HBC in scRNA-seq
FeaturePlot(rna, features = "Krtdap")
FeaturePlot(rna, features = "Krt6a")
FeaturePlot(rna, features = "Krt5")
FeaturePlot(rna, features = "Krt14")
FeaturePlot(rna, features = "Trp63")

# Markers for activated/resting HBC in scATAC-seq
FeaturePlot(atac, features = "Krtdap") + xlim(c(-5,7))
FeaturePlot(atac, features = "Krt6a") + xlim(c(-5,7))
FeaturePlot(atac, features = "Krt5") + xlim(c(-6,6))
FeaturePlot(atac, features = "Krt14") + xlim(c(-6,6))
FeaturePlot(atac, features = "Trp63") + xlim(c(-6,6))
```

# Assign sub-hybrid labels to RNA-seq

```{r}
ctHybridSub <- as.character(rna$celltype)
hlpid <- which(rna@reductions$umap@cell.embeddings[,1] > 2 & rna@reductions$umap@cell.embeddings[,2]>-1)
ctHybridSub[hlpid] <- NA
hlp2id <- which(rna@reductions$umap@cell.embeddings[,1] > 2 & rna@reductions$umap@cell.embeddings[,2]>-1 & rna@reductions$umap@cell.embeddings[,2]<4)
ctHybridSub[hlp2id] <- "hybrid1"
hlp1id <- which(rna@reductions$umap@cell.embeddings[,1] > 2 & rna@reductions$umap@cell.embeddings[,2]>4)
ctHybridSub[hlp1id] <- "hybrid2"
rna$ctHybridSub <- ctHybridSub


DimPlot(rna, group.by = "ctHybridSub", label = FALSE, repel = TRUE) +
  ggtitle("scRNA-seq") 

table(ctHybridSub)
```

## Explore hybrid1 markers from ATAC

```{r}

expressionBoxplot <- function(rna, gene){
  counts <- as.vector(rna@assays$RNA[gene,])
  df <- data.frame(counts=log1p(counts),
                   celltype=rna$ctHybridSub)
  ggplot(df, aes(x=celltype, y=counts)) +
    geom_boxplot() +
    xlab("Cell state") +
    ylab("Log(Expression + 1)") +
    theme_bw() +
    theme(axis.text.x = element_text(angle=45, vjust=0.5)) +
    ggtitle(gene)
}
```


```{r}
hybrid1Markers <- c("Muc5b", "Sult1c1")

FeaturePlot(rna, features = "Muc5b")
expressionBoxplot(rna, "Muc5b")

FeaturePlot(rna, features = "Sult1c1")
expressionBoxplot(rna, "Sult1c1")

```


## Explore hybrid2 markers from ATAC

Note that Pax7 seems to fuse with a FOX TF.. https://www.genecards.org/cgi-bin/carddisp.pl?gene=PAX7

```{r}
hybrid2Markers <- c("Pax3", "Pax7", "Pax9", "Dmrt2", "2610016A17Rik")

FeaturePlot(rna, features = "Pax3")
expressionBoxplot(rna, "Pax3")

FeaturePlot(rna, features = "Pax7")
expressionBoxplot(rna, "Pax7")

FeaturePlot(rna, features = "Pax9")
expressionBoxplot(rna, "Pax9")

FeaturePlot(rna, features = "Dmrt2")
expressionBoxplot(rna, "Dmrt2")

```

